---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


### Loading and preprocessing the data

```{r, echo=TRUE}
# Clear all objects
rm(list=ls())

knitr::opts_chunk$set(fig.path = "PA1_template_files/figure-html/")
  
if(file.exists("activity.zip")) {
        if(!exists("activityData")){
                activityData <- read.csv(unz("activity.zip","activity.csv"))
        }
}

# Let's prepare the dates for later processing
activityData$date <- as.Date(activityData$date, "%Y-%m-%d")

# Rrebuild the data frame to include days of the week and flagged as weekday or weekend for later processing
activityData <- data.frame(date=activityData$date, weekday=tolower(weekdays(activityData$date)), steps=activityData$steps, interval=activityData$interval)
activityData <- cbind(activityData, daytype=ifelse(activityData$weekday == "saturday" | activityData$weekday == "sunday", "weekend", "weekday"))
activityData <- data.frame(date=activityData$date, weekday=activityData$weekday, daytype=activityData$daytype, interval=activityData$interval, steps=activityData$steps)
```



### What is mean total number of steps taken per day?

* Calculate the total number of steps taken per day
```{r}
dailySteps <- aggregate(steps ~ date, activityData, sum)
```

* Make a histogram of the total number of steps taken each day
```{r}
hist(dailySteps$steps, breaks=seq(from=0, to=25000, by=1500), main = paste("Steps per day"), col="blue", xlab="Step Count")
```

* Calculate and report the mean and median of the total number of steps taken per day
```{r echo=TRUE}
meanSteps <- mean(dailySteps$steps)
medianSteps <- median(dailySteps$steps)
```

* Mean: `r meanSteps`
* Median:  `r medianSteps`



### What is the average daily activity pattern?

* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE}
# Calculate average steps for each of 5-minute interval during a 24-hour period
stepsInterval <- aggregate(steps ~ interval, activityData, mean)
plot(stepsInterval$interval,stepsInterval$steps, type="l", xlab="5 minute interval", ylab="Step count",main="Average Number of Steps by Interval")
```

* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
maxInterval <- stepsInterval[which.max(stepsInterval$steps),1]
```
The interval with the max number of steps across all days is `r maxInterval`



### Imputing missing values
* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r echo=TRUE}
NAs <- sum(is.na(activityData$steps))
```
The total number of missing values is `r NAs`

* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r echo=TRUE}
# I will be simply finding the averages to fill in the missing values
# Need to first identify where the na's are.
na_idx <- which(is.na(activityData$steps))

# Now get the means where the indexes (not yet replacing)
meanOfTheIndexes <- rep(mean(activityData$steps, na.rm=TRUE), times=length(na_idx))
```

* Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r echo=TRUE}
# For comparison sake showing the data prior to replacement of the NA's
head(activityData)
```

```{r echo=TRUE}
# Now we replace the empty values of steps using the index as a locator and apply with the mean indexes
activityData[na_idx, "steps"] <- meanOfTheIndexes
```

```{r echo=TRUE}
# Showing the activity data after replacement
head(activityData)
```

* Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r echo=TRUE}
hist(dailySteps$steps, breaks=seq(from=0, to=25000, by=1500), main = paste("Steps per day (NA's replaced by mean)"), col="red", xlab="Step Count")
```

What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r echo=TRUE}
meanSteps <- mean(dailySteps$steps)
medianSteps <- median(dailySteps$steps)
```
* Mean: `r meanSteps`
* Median:  `r medianSteps`

* Do these values differ from the estimates from the first part of the assignment? 
** Really no difference since we used the mean.  It just averaged out.

### Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE}
library(lattice)

mean_data <- aggregate(activityData$steps, by=list(activityData$daytype, activityData$weekday, activityData$interval), mean)

# Rename the attributes
names(mean_data) <- c("daytype", "weekday", "interval", "mean")

xyplot(mean ~ interval | daytype, mean_data, 
       type="l", layout=c(1,2), lwd=1, 
       main = "Avg steps by interval", xlab="Interval", ylab="Number of steps")
# It appears that during the weekend the steps ramp up earlier in the day.

```
